---
title: "HDS5105 Assignment 5"
output: html_document
date: "2024-12-02"
---

#Project Description
The app allows users to interact with and analyze the trial data (`DIG.csv`). Key features include:

- Interactive data visualization
- Statistical summaries
- etc...

#Project Structure
- `data/`: contains our dataset (`DIG.csv`)
- `app.R`: the main shiny application code file
- `other/`: codebook and any other misc files
- 'README.md': description of the repository

#Data
The app uses the `DIG.csv` file, which contains the trial data. A codebook for understanding the variables can be found in the `DIG_code_book.pdf`.

#Requirements
To run this app, you will need the following R packages:

- shiny
- ggplot2
- dplyr
- etc...

These can be installed using:
install.packages(c("shiny", "ggplot2", "dplyr", etc...))

Prepare DIG file

```{r}
DIG.df <- read.csv("data/DIG.csv")
DIG_sub.df <- DIG.df %>%
  select(ID, TRTMT, AGE, SEX, BMI, KLEVEL, CREAT, DIABP, SYSBP, 
         HYPERTEN, CVD, WHF, DIG, HOSP, HOSPDAYS, DEATH, DEATHDAY)
DIG_sub.df <- DIG_sub.df %>%
  mutate(
    ID = as.factor(ID),
    TRTMT = as.factor(TRTMT),
    AGE = as.numeric(AGE),
    SEX = as.factor(SEX),
    BMI = as.numeric(BMI),
    KLEVEL = as.numeric(KLEVEL),
    CREAT = as.numeric(CREAT),
    DIABP = as.numeric(DIABP),
    SYSBP = as.numeric(SYSBP),
    HYPERTEN = as.factor(HYPERTEN),
    CVD = as.factor(CVD),
    WHF = as.factor(WHF),
    DIG = as.factor(DIG),
    HOSP = as.factor(HOSP),
    HOSPDAYS = as.numeric(HOSPDAYS),
    DEATH = as.factor(DEATH),
    DEATHDAY = as.numeric(DEATHDAY)
  )
DIG_sub.df <- DIG_sub.df %>%
  mutate(
    TRTMT = fct_recode(TRTMT, Placebo = "0", Treatment = "1"),
    SEX = fct_recode(SEX, Male = "1", Female = "2"),
    HYPERTEN = fct_recode(HYPERTEN, No = "0", Yes = "1"),
    CVD = fct_recode(CVD, No = "0", Yes = "1"),
    WHF = fct_recode(WHF, No = "0", Yes = "1"),
    DIG = fct_recode(DIG, No = "0", Yes = "1"),
    HOSP = fct_recode(HOSP, No = "0", Yes = "1"),
    DEATH = fct_recode(DEATH, Alive = "0", Death = "1")
  )
```

Shiny

```{r}
if (!require("shiny")) install.packages("shiny")
if (!require("tidyverse")) install.packages("tidyverse")
if (!require("DynNom")) install.packages("DynNom")
if (!require("shinydashboard")) install.packages("shinydashboard")
library(tidyverse)
library(shiny)



ui <- fluidPage(
  titlePanel("DIG Trial Data Exploration"),
  sidebarLayout(
    sidebarPanel(
      selectInput(inputId = "TRTMT", label = "Treatment Group:", choices = c("Placebo", "Treatment"), multiple = FALSE),
      radioButtons(inputId = "SEX", label = "Select Sex:", choices = c("Male", "Female"), inline = T)
    ),
    mainPanel(
      #outputs here
    )
  )
)

server <- function(input, output) {
  #server workings here
}

shinyApp(ui = ui, server = server)

```