---
title: "HDS5105 Assignment 5"
output: html_document
date: "2024-12-22"
---

#Project Description
The app allows users to interact with and analyze the trial data (`DIG.csv`). Key features include:

- Interactive data visualization
- Statistical summaries
- etc...

#Project Structure
- `data/`: contains our dataset (`DIG.csv`)
- `app.R`: the main shiny application code file
- `other/`: codebook and any other misc files
- 'README.md': description of the repository

#Data
The app uses the `DIG.csv` file, which contains the trial data. A codebook for understanding the variables can be found in the `DIG_code_book.pdf`.

#Requirements
To run this app, you will need the following R packages:

- shiny
- ggplot2
- dplyr
- etc...

These can be installed using:
install.packages(c("shiny", "ggplot2", "dplyr", etc...))

Prepare DIG file


Shiny

```{r}
install.packages("table1")
install.packages("dplyr")
install.packages("ggplot2")
install.packages("forcats")
install.packages("shinydashboard")
library(forcats)
library(readr)
library(dplyr)
library(table1)
library(ggplot2)
library(tidyverse)
library(shiny)
library(shinydashboard)
```

```{r}
DIG.df <- read.csv("data/DIG.csv") %>%
  mutate(
    TRTMT = fct_recode(as.factor(TRTMT), Placebo = "0", Treatment = "1"),
    SEX = fct_recode(as.factor(SEX), Male = "1", Female = "2"),
    HYPERTEN = fct_recode(as.factor(HYPERTEN), No = "0", Yes = "1"),
    CVD = fct_recode(as.factor(CVD), No = "0", Yes = "1"),
    WHF = fct_recode(as.factor(WHF), No = "0", Yes = "1"),
    HOSP = fct_recode(as.factor(HOSP), No = "0", Yes = "1"),
    DEATH = fct_recode(as.factor(DEATH), Alive = "0", Death = "1"),
    AGE = as.numeric(AGE),
    BMI = as.numeric(BMI)
  )

ui <- dashboardPage(
  dashboardHeader(title = tags$div(
      "Digitalis Investigation Group",
      style = "font-size: 24px; color: White; text-align: center; width: 100%;"),
    titleWidth = "100%"),
  dashboardSidebar(checkboxGroupInput(inputId = "sex", label = "Select Sex:", choices = c("Male", "Female"), selected = c("Male", "Female")),
    sliderInput("age_range", "Select Age Range:", min = min(DIG.df$AGE, na.rm = TRUE), max = max(DIG.df$AGE, na.rm = TRUE), value = c(50, 80)),
    sliderInput("bmi_range", "Select BMI Range:", min = min(DIG.df$BMI, na.rm = TRUE), max = max(DIG.df$BMI, na.rm = TRUE), value = c(20, 30)),
    checkboxGroupInput(inputId = "hypertension", label = "Hypertension:", choices = c("Yes", "No"), selected = c("Yes", "No")),
    checkboxGroupInput(inputId = "cvd", label = "Cardiovascular Disease:", choices = c("Yes", "No"), selected = c("Yes", "No")),
    checkboxGroupInput(inputId = "hospitalization", label = "Hospitalization:", choices = c("Yes", "No"), selected = c("Yes", "No")),
    checkboxGroupInput(inputId = "whf", label = "Worsening Heart Failure:", choices = c("Yes", "No"), selected = c("Yes", "No"))
  ),
  dashboardBody(tabBox(width = 12,id = "tabs",
      tabPanel("Mortality Rate Plot",
        fluidRow(
          box(width = 12,title = div("Mortality Rate by Treatment Group",style = "text-align: center; line-height: 30px; color: white;"),
            collapsible = TRUE,status = "success",solidHeader = TRUE,plotOutput("plot1"))
               )),
      tabPanel("Filtered Data",
        fluidRow(
          box(width = 12,title = tags$div("Filtered Data Table",style = "text-align: center; line-height: 30px; color: white;"),
            collapsible = TRUE,status = "success",solidHeader = TRUE,dataTableOutput("table1"))
               ))
    )
  ),
   skin = "green")
server <- function(input, output) {
  filtered_data <- reactive({
    DIG.df %>%
      filter(SEX %in% input$sex) %>%
      filter(AGE >= input$age_range[1] & AGE <= input$age_range[2]) %>%
      filter(BMI >= input$bmi_range[1] & BMI <= input$bmi_range[2]) %>%
      filter(HYPERTEN %in% input$hypertension) %>%
      filter(CVD %in% input$cvd) %>%
      filter(HOSP %in% input$hospitalization) %>%
      filter(WHF %in% input$whf)
  })
  output$plot1 <- renderPlot({
    filtered_data() %>%
      group_by(TRTMT) %>%
      summarise(Mortality_Rate = mean(as.numeric(DEATH == "Death"), na.rm = TRUE)) %>%
      ggplot(aes(x = TRTMT, y = Mortality_Rate, fill = TRTMT)) +
      geom_bar(stat = "identity", position = "dodge", alpha = 1, width = 0.5) +
      geom_text(aes(label = scales::percent(Mortality_Rate, accuracy = 0.1)), 
                vjust = -0.3, size = 8) +
       scale_fill_manual(values = c("black", "skyblue")) +
      theme_classic() +
      labs(x = "Treatment Group",y = "Mortality Rate",fill = "Treatment Group") +
      theme(
        plot.title = element_text(color = "black", size = 20, face = "bold")
      ) +
      scale_y_continuous(labels = scales::percent, limits = c(0, 1))
  })
  
output$table1 <- renderDataTable({
    filtered_data() %>%
      select(ID, TRTMT, AGE, SEX, BMI, KLEVEL, CREAT, DIABP, SYSBP, 
         HYPERTEN, CVD, WHF, DIG, HOSP, HOSPDAYS, DEATH, DEATHDAY)
})
}
shinyApp(ui = ui, server)
```

Not sure how you guys would like it for the format. For now I will put my work in the chunk below. (KY)

```{r}
library(tidyverse)
library(shiny)
library(ggplot2)

# Purpose - Shinny App to show prognostic factors effect on death rate

ui <- fluidPage(
  titlePanel("Prognostic Effect on Death Rate by Group"),
  
  sidebarLayout(
    sidebarPanel(
        sliderInput(inputId = "AGE", label = "Select Age", min = 0, max = 100, value = c(0,100), step = 1,animate = TRUE),
        selectInput(inputId = "SEX", label = "Select Sex", choices = c("Male","Female"), multiple = T, selected = c("Male", "Female")),
        sliderInput(inputId = "BMI", label = "Select BMI", min = 10, max = 65, value = c(10,65)),
        sliderInput(inputId = "KLEVEL", label = "Select Potassium Level", min = 0, max = 450, value = c(0,450)),
        sliderInput(inputId = "CREAT", label = "Select Creatinine Level", min = 0, max = 4, value = c(0,4)),
        selectInput(inputId = "HYPERTEN", label = "Select Hypertension History", choices = c("Yes","No"), multiple = T, selected = c("Yes", "No")),
        selectInput(inputId = "CVD", label = "Select CVD History", choices = c("Yes","No"), multiple = T, selected = c("Yes", "No")),
        selectInput(inputId = "WHF", label = "Select Worsen HF History", choices = c("Yes","No"), multiple = T, selected = c("Yes", "No")),
        selectInput(inputId = "DIG", label = "Select Digoxin History", choices = c("Yes","No"), multiple = T, selected = c("Yes", "No"))
    ),
    
    mainPanel(
    plotOutput("plot1"),
    dataTableOutput("table1")
    )
  )
  
)

server <- function(input, output) {
  
  DIG_sub.df2 <- reactive({
    DIG_sub.df %>%
      filter(AGE >= input$AGE[1] & AGE <= input$AGE[2]) %>%
      filter(SEX == input$SEX) %>%
      filter(BMI >= input$BMI[1] & BMI <= input$BMI[2]) %>%
      filter(KLEVEL >= input$KLEVEL[1] & KLEVEL <= input$KLEVEL[2]) %>%
      filter(CREAT >= input$CREAT[1] & CREAT <= input$CREAT[2]) %>%
      filter(HYPERTEN == input$HYPERTEN) %>%
      filter(CVD == input$CVD) %>%
      filter(WHF == input$WHF) %>%
      filter(DIG == input$DIG)
  })
   
  output$plot1 <- renderPlot({ 
      DIG_sub.df2() %>%
      group_by(TRTMT) %>%
      summarise(Death_rate = sum((DEATH == "Death") / n())) %>%
      ggplot(aes(x = TRTMT, y = Death_rate, fill = TRTMT)) + geom_bar (stat = "identity")
    })
  
  output$table1 <- renderDataTable({ 
    DIG_sub.df2()
    })
  
}

shinyApp(ui, server)
```

