---
title: "HDS5105 Assignment 5"
output: html_document
date: "2024-12-23"
---


Shiny app for DIG trial data exploration

This repository contains a Shiny app designed for exploring and visualizing data from the Digitalis Investigation Group (DIG) Trial.
It allows users to interact with and analyze the trial data (`DIG.csv`) via interactive data visualization and statistical summaries.

- `data/`: contains the dataset (`DIG.csv`)
- `HDS5105-Assignment-5.Rmd`: the main shiny application code file
- `other/`: codebook and any other misc files
- 'README.md': description of the repository

The app uses the `DIG.csv` file, which contains the trial data. A codebook for understanding the variables can be found in the `DIG_code_book.pdf`.

To run this app, you will need the following R packages:

- shiny
- ggplot2
- dplyr
- forcats
- readr
- table1
- tidyverse
- shinydashboard
- plotly
- Stat2Data

```{r}
install.packages("table1")
install.packages("dplyr")
install.packages("ggplot2")
install.packages("forcats")
install.packages("shinydashboard")
install.packages("plotly")
install.packages("Stat2Data")
library(plotly)
library(forcats)
library(readr)
library(dplyr)
library(table1)
library(ggplot2)
library(tidyverse)
library(shiny)
library(shinydashboard)
library(Stat2Data)
```

#Prepare DIG file

```{r}
DIG.df <- read.csv("data/DIG.csv") %>%
  mutate(
    TRTMT = fct_recode(as.factor(TRTMT), Placebo = "0", Treatment = "1"),
    SEX = fct_recode(as.factor(SEX), Male = "1", Female = "2"),
    HYPERTEN = fct_recode(as.factor(HYPERTEN), No = "0", Yes = "1"),
    CVD = fct_recode(as.factor(CVD), No = "0", Yes = "1"),
    WHF = fct_recode(as.factor(WHF), No = "0", Yes = "1"),
    HOSP = fct_recode(as.factor(HOSP), No = "0", Yes = "1"),
    DEATH = fct_recode(as.factor(DEATH), Alive = "0", Death = "1"),
    AGE = as.numeric(AGE),
    BMI = as.numeric(BMI)
  )
```

#Shiny

```{r}
ui <- dashboardPage(
  
  #The Header
  dashboardHeader(title = tags$div("Digitalis Investigation Group",
      style = "font-size: 24px; color: white; text-align: center; width: 100%;"),
    titleWidth = "100%"),
  
  #The Sidebar
  dashboardSidebar(
    checkboxGroupInput(inputId = "sex", label = "Select Sex:", choices = c("Male", "Female"), selected = c("Male", "Female")),
    sliderInput("age_range", "Select Age Range:", min = min(DIG.df$AGE, na.rm = TRUE), 
                max = max(DIG.df$AGE, na.rm = TRUE), 
                value = c(min(DIG.df$AGE, na.rm = TRUE), max(DIG.df$AGE, na.rm = TRUE))),
    sliderInput("bmi_range", "Select BMI Range:", min = min(DIG.df$BMI, na.rm = TRUE), 
                max = max(DIG.df$BMI, na.rm = TRUE), value = c(min(DIG.df$BMI, na.rm = TRUE), max(DIG.df$BMI, na.rm = TRUE))),
    checkboxGroupInput(inputId = "hypertension", label = "Hypertension:", choices = c("Yes", "No"), selected = c("Yes", "No")),
    checkboxGroupInput(inputId = "cvd", label = "Cardiovascular Disease:", choices = c("Yes", "No"), selected = c("Yes", "No")),
    checkboxGroupInput(inputId = "hospitalization", label = "Hospitalization:", choices = c("Yes", "No"), selected = c("Yes", "No")),
    checkboxGroupInput(inputId = "whf", label = "Worsening Heart Failure:", choices = c("Yes", "No"), selected = c("Yes", "No"))
  ),
  
  #Dash Body
  dashboardBody(tabBox(width = 12,id = "tabs",
                       
      #DIG Info tab
      tabPanel("DIG Information",
        fluidRow(
          box(width = 12,title = tags$div("DIG Trial Information",style = "text-align: center; font-weight: bold; color: white;"),
            collapsible = TRUE,status = "success",solidHeader = TRUE,
            tags$div(
              style = "padding: 15px;",
              tags$p(
                "The DIG (Digitalis Investigation Group) Trial was a randomized, double-blind, multicenter trial with more than 300 centers in the United States and Canada participating. 
                The purpose of the trial was to examine the safety and efficacy of Digoxin in treating patients with congestive heart failure in sinus rhythm."
              ),
              tags$p(
                "Digitalis was introduced clinically more than 200 years ago and has since become a commonly prescribed medication for the treatment of heart failure; however, there was considerable uncertainty surrounding its safety and efficacy."
              ),
              tags$p(
                "Small trials indicated that Digoxin alleviated some of the symptoms of heart failure, prolonged exercise tolerance, and generally improved the quality of patients’ lives. 
                Unfortunately, these trials were generally small and although they did focus on the effect of treatment on patients’ relief from heart failure symptoms and quality of life, they failed to address the effect of treatment on cardiovascular outcomes."
              ),
              tags$p(
                "Questions about the safety of Digoxin were also a concern. Digoxin toxicity is uncommon in small trials with careful surveillance; however, the long-term effects of therapeutic levels of Digoxin were less clear."
              )
            )
          )
        )
      ),
      
      #Scatter Plot tab
      tabPanel(
        "Scatter Plot",
        fluidRow(
          box(width = 12,title = tags$div("Scatter Plot",
              style = "text-align: center; color: white;"
            ),
            collapsible = TRUE,status = "success",solidHeader = TRUE,
            selectInput("y_var", "Select Y-Axis Variable:", 
                        choices = c("BMI", "SYSBP", "DIABP", "KLEVEL"), selected = "BMI"),
            plotlyOutput("scatter_plot")
          )
        )
      ),
      
      tabPanel(
        "Mortality Rate",
        fluidRow(
          box(width = 12,title = tags$div("Mortality Rate by Treatment Group",
              style = "text-align: center; line-height: 30px; color: white;"
            ),
            collapsible = TRUE,status = "success",solidHeader = TRUE,plotlyOutput("plot1")
          )
        )
      ),
      
      #Data tab
      tabPanel(
        "Data",
        fluidRow(
          box(width = 12,title = tags$div("Data Table",
              style = "text-align: center; line-height: 30px; color: white;"
            ),
            collapsible = TRUE,status = "success",solidHeader = TRUE,dataTableOutput("table1")
          )
        )
      ),
      
      #OR tab
      tabPanel(
        "Odds Ration",
        fluidRow(
          box(width = 12,title = tags$div("Odds Ration and Confidence Interval",
              style = "text-align: center; line-height: 30px; color: white;"
            ),
            collapsible = TRUE,status = "success",solidHeader = TRUE,plotlyOutput("OR_plot")
          )
        )
      ),
      
      #Death curve tab
      tabPanel(
        "Death curve",
        fluidRow(
          box(width = 12,title = tags$div("Cumulative Death Portion by Day",
              style = "text-align: center; line-height: 30px; color: white;"
            ),
            collapsible = TRUE,status = "success",solidHeader = TRUE,plotlyOutput("death_plot")
          )
        )
      ),
      
    )
  ),
  
  skin = "green" #Skin
)

server <- function(input, output) {
  
  #Prepare data
  filtered_data <- reactive({
    DIG.df %>%
      filter(SEX %in% input$sex) %>%
      filter(AGE >= input$age_range[1] & AGE <= input$age_range[2]) %>%
      filter(BMI >= input$bmi_range[1] & BMI <= input$bmi_range[2]) %>%
      filter(HYPERTEN %in% input$hypertension) %>%
      filter(CVD %in% input$cvd) %>%
      filter(HOSP %in% input$hospitalization) %>%
      filter(WHF %in% input$whf)
  })
  
  # Scatter plot
  output$scatter_plot <- renderPlotly({
    data <- filtered_data()
    p <- ggplot(data, aes(x = AGE, y = .data[[input$y_var]], color = SEX)) +
      geom_point(alpha = 0.7, size = 1.5) +
      scale_color_manual(values = c("Male" = "black", "Female" = "pink")) +
      theme_minimal() +
      labs(x = "Age",y = input$y_var,color = "Sex",
        title = paste("Scatter Plot of", input$y_var, "vs Age")
      )
    
    ggplotly(p)
  })
  
  # Mortality rate
  output$plot1 <- renderPlotly({
    data <- filtered_data() %>%
      group_by(TRTMT) %>%
      summarise(
        Deaths = sum(DEATH == "Death", na.rm = TRUE),
        Mortality_Rate = mean(as.numeric(DEATH == "Death"), na.rm = TRUE)
      )
    
    p <- ggplot(data, aes(x = TRTMT, y = Mortality_Rate, fill = TRTMT, text = paste(
      "Treatment Group: ", TRTMT, "<br>",
      "Deaths: ", Deaths, "<br>",
      "Mortality Rate: ", scales::percent(Mortality_Rate, accuracy = 0.1)
    ))) +
      geom_bar(stat = "identity", position = "dodge", alpha = 1, width = 0.5) +
      scale_fill_manual(values = c("black", "skyblue")) +
      theme_classic() +
      labs(x = "Treatment Group",y = "Mortality Rate (%)",fill = "Treatment Group"
      ) +
      scale_y_continuous(labels = scales::percent, limits = c(0, 1))
    
    ggplotly(p, tooltip = "text")
  })
  
  #Data table
  output$table1 <- renderDataTable({
    filtered_data() %>%
      select(
        ID, TRTMT, AGE, SEX, BMI, KLEVEL, CREAT, DIABP, SYSBP,
        HYPERTEN, CVD, WHF, DIG, HOSP, HOSPDAYS, DEATH, DEATHDAY
      )
  }, options = list(pageLength = 10,lengthMenu = c(5, 10, 25, 50),scrollX = TRUE,autoWidth = TRUE
  ))
  
  #OR plot
  output$OR_plot <- renderPlotly({
    df <- filtered_data()
    #fit glm model
    model = glm(DEATH ~ TRTMT, family = binomial, data = df)
    #tablelize coefficients
    or_data <- data.frame(
      DEATH = "TRTMT VS PLACEBO",
      Middle = exp(model$coefficients[2]),
      Lower = exp(confint(model)[2,1]),     
      Upper = exp(confint(model)[2,2])
      )
    #ggplot
    ggplot(or_data, aes(x = DEATH, y = Middle)) +
      geom_errorbar(aes(ymin = Lower, ymax = Upper), width = 0.2, color = "blue", size = 1.2) +
      geom_point(size = 4, color = "black") +
      labs(
        title = "Odds Ratio of Death, Treatment vs Placebo, With Confidence Interval",
        x = NULL,
        y = "Odds Ratio"
        ) +
      theme_minimal()
  })
  
  #Death curve
  output$death_plot <- renderPlotly({
    filtered_data() %>%
  ggplot(aes(x = DEATHDAY, color = TRTMT)) +
  stat_ecdf(geom = "step", alpha=0.8, size=1) + 
  scale_color_manual(values = c("red", "darkgreen")) +
  theme_minimal() +
  labs(x = "Death Day",y = "Cumulative Death Rate (%)",fill = "Treatment Group")
  })
}

shinyApp(ui = ui, server = server)
```

